MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-08-02 MAESTROeX_I/equations.tex
MZ read through

By comparing the momentum equation (\ref{eq:momentum}) to the momentum equation used in equation (2) in Paper V, we
note that we are using a formulation that enforces conservation of total energy in the
low Mach number system in the absence of external heating or viscous terms \citep{kleinpauluis,Vasil2013}.
We have previously validated this approach in modeling sub-Chandrasekhar white dwarfs using MAESTRO \citep{subChandra_II}.
We also define a one-dimensional base state density, $\rho_0 = \rho_0(r,t)$, that represents the lateral average (see Section \ref{Sec:Spatial}) of $\rho$ and is in hydrostatic equilibrium with $p_0$, i.e.,
\begin{equation}
\nabla p_0 = -\rho_0 g\eb_r, \label{eq:HSE}
\end{equation}
where $g=g(r,t)$ is the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the outward radial direction.
Here $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho) |_s$ (evaluated with entropy, $s$, constant).
Thermal diffusion is not discussed in this paper, but we have previously described the modifications to the original algorithm required
for implicit thermal diffusion in \cite{XRB_I}; inclusion of these effects in the new algorithm presented here is straightforward.

>>>>

By comparing the momentum equation (\ref{eq:momentum}) to the momentum equation used in equation (2) in Paper V, we
note that we are using a formulation that enforces conservation of total energy in the
low Mach number system in the absence of external heating or viscous terms \citep{kleinpauluis,Vasil2013}.
We have previously validated this approach in modeling sub-Chandrasekhar white dwarfs using MAESTRO \citep{subChandra_II}.
We also define a one-dimensional base state density, $\rho_0 = \rho_0(r,t)$, that represents the lateral average (see Section \ref{Sec:Spatial}) of $\rho$ and is in hydrostatic equilibrium with $p_0$, i.e.,
\begin{equation}
\nabla p_0 = -\rho_0 g\eb_r, \label{eq:HSE}
\end{equation}
where $g=g(r,t)$ is the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the outward radial direction.
Here $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho) |_s$ (evaluated with entropy, $s$, constant).  We explored the effect of
replacing $\Gamma_1$ with $\gammaonebar$ as well as a correction term in paper III. \MarginPar{MZ added this sentence--good to remember this}
Thermal diffusion is not discussed in this paper, but we have previously described the modifications to the original algorithm required
for implicit thermal diffusion in \cite{XRB_I}; inclusion of these effects in the new algorithm presented here is straightforward.

### 2019-08-02 MAESTROeX_I/intro.tex
MZ read through

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
MAESTRO is a block-structured, Cartesian grid finite-volume, adaptive mesh refinement (AMR)
code that has been successfully used for many years for a number of applications, detailed below.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for
flows where the Mach number is small ($\sim 0.1$ or smaller).
Furthermore, the low Mach number model in MAESTRO is specifically designed for, but not limited
to, astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized
regions of a star.
The numerical methodology relies on an explicit Godunov approach for advection, a stiff ODE
solver for reactions (VODE) \citep{vode}, and multigrid-based linear solvers for the
pressure-projection steps.  Thus, the time step is limited by an advective CFL constraint based on the 
fluid velocity, not the sound speed.
Central to the algorithm are time-varying, one-dimensional stratified background (or base) state
density and pressure fields that are held in hydrostatic equilibrium.
The base state density couples to the full state solution through buoyancy terms in the momentum equation,
and the base state pressure couples to the full state solution by constraining the evolution of the
thermodynamic variables to match this pressure.
The time-advancement strategy uses Strang splitting to integrate the thermodynamic variables, a
second-order projection method to integrate the velocity subject to a divergence constraint,
and a velocity splitting scheme that uses a radially-averaged velocity to hydrodynamically evolve the base state.\MarginPar{AJN changed this sentence.}
The original MAESTRO code was developed in the pure-Fortran 90 FBoxLib software framework, whereas
MAESTROeX is developed in the C++/F90 AMReX framework \citep{AMReX,AMReX_JOSS}.

@@

In this paper, we present new algorithmic methodology that improves upon Paper V in a number of ways.
First, the overall temporal algorithm has been greatly simplified without compromising second-order accuracy.
The key design decisions were to eliminate the splitting of the velocity into average and perturbational components,
and also to replace the hydrodynamic evolution of the base state with a predictor-corrector approach.
Not only does this greatly simplify the dynamics of the base
state, but this treatment is more amenable to higher-order multiphysics coupling strategies
based on method-of-lines integration.
In particular, schemes based on deferred corrections \citep{dutt2000spectral} have been used to generate 
high-order temporal integrators for problems of reactive flow and low Mach number combustion \citep{pazner2016high,nonaka2018conservative}.
Second, we explore the effects of alternative spatial mapping routines for coupling the base state and the Cartesian grid state for spherical problems.
Finally, we examine the performance of our new MAESTROeX implementation in the new C++/F90 AMReX public software library \citep{AMReX,AMReX_JOSS}.
MAESTROeX uses MPI+OpenMP parallelism and scales well to over 10,000 MPI processes, with each MPI process supporting tens of threads.
The resulting code is publicly available on GitHub ({\tt https://github.com/AMReX-Astro/MAESTROeX}),
uses the Starkiller-Astro microphysics libraries \citep{starkiller} ({\tt https://github.com/starkiller-astro}) ,
as well as AMReX ({\tt https://github.com/AMReX-Codes/amrex}).

>>>>

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
MAESTRO is a block-structured, Cartesian grid finite-volume, adaptive mesh refinement (AMR)
code that has been successfully used for many years for a number of applications, detailed below.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for
flows where the Mach number is small ($\sim 0.1$ or smaller).
Furthermore, the low Mach number model in MAESTRO is specifically designed for, but not limited
to, astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized
regions of a star.
The numerical methodology relies on an explicit Godunov approach for advection, a stiff ODE
solver for reactions (VODE, \citealt{vode}), and multigrid-based linear solvers for the
pressure-projection steps.  Thus, the time step is limited by an advective CFL constraint based on the 
fluid velocity, not the sound speed.
Central to the algorithm are time-varying, one-dimensional stratified background (or base) state
density and pressure fields that are held in hydrostatic equilibrium.
The base state density couples to the full state solution through buoyancy terms in the momentum equation,
and the base state pressure couples to the full state solution by constraining the evolution of the
thermodynamic variables to match this pressure.
The time-advancement strategy uses Strang splitting to integrate the thermodynamic variables, a
second-order projection method to integrate the velocity subject to a divergence constraint,
and a velocity splitting scheme that uses a radially-averaged velocity to hydrodynamically evolve the base state.\MarginPar{AJN changed this sentence.}
The original MAESTRO code was developed in the pure-Fortran 90 FBoxLib software framework, whereas
MAESTROeX is developed in the C++/F90 AMReX framework \citep{AMReX,AMReX_JOSS}.

@@

In this paper, we present new algorithmic methodology that improves upon Paper V in a number of ways.
First, the overall temporal algorithm has been greatly simplified without compromising second-order accuracy.
The key design decisions were to eliminate the splitting of the velocity into average and perturbational components,
and also to replace the hydrodynamic evolution of the base state with a predictor-corrector approach.
Not only does this greatly simplify the dynamics of the base
state, but this treatment is more amenable to higher-order multiphysics coupling strategies
based on method-of-lines integration.
In particular, schemes based on deferred corrections \citep{dutt2000spectral} have been used to generate 
high-order temporal integrators for problems of reactive flow and low Mach number combustion \citep{pazner2016high,nonaka2018conservative}.
Second, we explore the effects of alternative spatial mapping routines for coupling the base state and the Cartesian grid state for spherical problems.
Finally, we examine the performance of our new MAESTROeX implementation in the new C++/F90 AMReX public software library \citep{AMReX,AMReX_JOSS}.
MAESTROeX uses MPI+OpenMP parallelism and scales well to over 10,000 MPI processes, with each MPI process supporting tens of threads.
The resulting code is publicly available on GitHub (\url{https://github.com/AMReX-Astro/MAESTROeX}),
uses the Starkiller-Astro microphysics libraries (\citealt{starkiller}, \url{https://github.com/starkiller-astro}) ,
as well as AMReX (\url{https://github.com/AMReX-Codes/amrex}).
